Semi-supervised low-rank logistic regression for high-dimensional neuroimaging data

### 2015-06-01 paper_nips/paper.tex
A pass on the end of the paper

The labeled and unlabeled data were fed into a linear statistical model
composed of autoencoder and low-rank logistic regression.

\paragraph{Linear autoencoder.}
The affine autoencoder takes the input 
$\mathbf{x}$ and projects it into coordinates of a
latent representation $\mathbf{z}$
by

@@

\begin{equation}
  \mathbb E\left[{\mathcal{L_{AE}}}(\mathbf{x})\right] = %
  \mathbb E\left[\| \mathbf{x} - f_{AE}(\mathbf{x}) \|^2\right],
\end{equation}
with \(f_{AE}(\mathbf x) = %
\mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}\)
the autoencoder.
% XXX Michael: I don't know to what extent the following makes sense or is
% XXX correct, so I am commenting it
% XXX Danilo: I found this in Bengio's papers somewhere
This reconstruction error criterion equates with
maximizing a lower bound on the mutual information between
input and the learned representation.
Here we choose $\mathbf{W_0}$ and $\mathbf{W_1}$ to be tied, i.e.
$\mathbf{W_0} = \mathbf{W_1^T}$. In particular, this means that the weights
learned are forced to take a two-fold function: That of signal 
\textit{analysis} and that of signal \textit{synthesis}.
The first layer \textit{analyzes}
the data in order to obtain the cleanest latent representation,
while the second
layer represents building blocks from which to \textit{synthesize} the data
using the latent activations.
Tying these processes together makes the analysis
layer interpretable and pulls all non-zero singular values towards 1. 
Nonlinearities were not applied to the
activations in the first layer.

@@

\paragraph{Optimization.}
The common objective was optimized
in the neuroimaging data
by gradient descent in the parameters
of the 
semi-supervised low-rank logistic regression (SS-LR LogReg).
The required gradients are easily obtained by using the chain rule to
backpropagate error derivatives through the linear network.
As solver, we chose \textit{rmsprop} \cite{rmsprop}.
%, a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by rmsprop
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. The bias parameters were initalized to zero.
With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch

@@

% XXX Michael: Probably remove the explanations following and just state values
where $0 < \rho < 1$ constitutes the decay rate. $\rho$ was set to
$0.9$ to deemphasize the magnitude of the gradient.
Further, $\alpha$ is the learning rate and $\epsilon$ a global damping factor.
The hyper-parameter $\alpha$ was set to $0.00001$ by prior studies and $\epsilon$ was set to $10^{-6}$.
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
rmsprop converged faster and with higher generalization performance.

@@

\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}
We first tested for an advantage of combined unsupervised decomposition
and supervised classification learning.
We benchmarked against
performing data reduction on the (unlabeled)
first half of the HCP data by ICA and SPCA ($n=5, 20, 50, 100$ components)
and learning classification models in the (labeled) second half
by ordinary logistic regression.
%
ICA performed iterative blind source separation
by a parallel FASTICA implementation (200 maximum iterations,
per-iteration tolerance of 0.0001,
initialized by random mixing matrix, whitening).
SPCA separated the BOLD signals into
network components with few regions by
a regression-type optimization problem constrained by
$\ell_1$-penalty
(no orthogonality assumptions, 1000 maximum iterations,
per-iteration tolerance of 1 * 10\textsuperscript{-8}, sparsity
alpha=1).
%
The second half of the data was projected onto the
latent components discovered in first data half.
The ensuing component loadings were submitted to ordinary
logistic regression
(one hidden layer, $\ell_1=0.1$, $\ell_2=0.1$, 500 maximum iterations).
%
This serail two-step approach was
compared against SS-LR LogReg
(two hidden layers, $\ell_1=0.1$, $\ell_2=0.1$,
500 maximum iterations, $\lambda=0.75$).
Importantly, all trained classification models were tested
on a large, unseen test set (20\% of data) in the presented analyses.
%
Across choices for $n$, SS-LR LogReg
achieved more than 96\% out-of-sample accuracy, whereas
supervised learning based on ICA and SPCA loadings
ranged from 37,53\% to 87,28\%
and
32,19\% to 83,97\%, respectively (Table \ref{table_one}).
%
These explorations attest to the advantage of directly searching for
classification-relevant structure in the fMRI data,
rather than solving supervised and unsupervised problems independently.
This effect was particularly pronounced when assuming few hidden dimensions.

\begin{table}[h]
 \centering
 \resizebox{0.5\textwidth}{!}{%
 \begin{tabular}{lccc}
 \hline
 \textit{n}               & ICA + LogReg & SPCA + LogReg & SS-LR LogReg \\ \hline
 \multicolumn{1}{l|}{5}   & 37,53 \%     & 32,19 \%      & 96,50 \%     \\
 \multicolumn{1}{l|}{20}  & 80,98 \%     & 78,15 \%      & 97,33 \%     \\
 \multicolumn{1}{l|}{50}  & 84,19 \%     & 83,97 \%      & 97,69 \%     \\
 \multicolumn{1}{l|}{100} & 87,28 \%     & 82,19 \%      & 97,80 \%     \\ \hline
 \end{tabular}
 }
 \vspace{-0.2cm}
 \caption{Serial versus parallel dimensionality reduction and classification}
 \label{table_one}
 \end{table}

@@

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\begin{tabular}{l|cccc|cccc|cccc|cccc|}
  \hline
                            & \multicolumn{4}{c|}{\textit{n = 5}}                                                                                                            & \multicolumn{4}{c|}{\textit{n = 20}}                                                                                                                   & \multicolumn{4}{c}{\textit{n = 100}}                                                       \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.5$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1$} & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.5$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1$}   & $\lambda=0.25$   & $\lambda=0.5$    & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1$}    \\ \hline
  Out-of-sample\\accuracy   & \textit{88,90\%}                   & 95,13\%                            & \textbf{96,49\%}                  & 95,72\%                          & 97,44\%                              & \textbf{97,80\%}                    & 97,33\%                              & \textit{97,32\%}                   & 97,21\%          & \textit{97,03\%} & \textbf{97,80\%} & 97,38\%                            \\
  Precision (mean)          & \textit{86,98\%}                   & 94,86\%                            & \textbf{96,28\%}                  & 95,36\%                          & \textbf{97,38\%}                     & 97,06\%                             & \textit{96,98\%}                     & 97,02\%                            & 96,90\%          & \textit{96,47\%} & \textbf{97,50\%} & 96,91\%                            \\
  Recall (mean)             & \textit{88,27\%}                   & 95,19\%                            & \textbf{96,57\%}                  & 95,70\%                          & \textbf{97,51\%}                     & 97,50\%                             & \textit{97,35\%}                     & 97,36\%                            & 97,20\%          & \textit{97,15\%} & \textbf{97,88\%} & 97,37\%                            \\
  F1 score (mean)           & \textit{86,64\%}                   & 94,89\%                            & \textbf{96,38\%}                  & 95,43\%                          & \textbf{97,42\%}                     & 97,22\%                             & \textit{97,11\%}                     & 97,13\%                            & 97,00\%          & \textit{96,71\%} & \textbf{97,65\%} & 97,07\%                            \\ \hline
  Out-of-dataset \\accuracy & 60,83\%                            & \textit{54,30\%}                   & 60,69\%                           & \textbf{62,92\%}                 & 79,72\%                              & \textbf{81,94\%} & 79,72\%          & \textit{79,44\%}                     & \textbf{82,08\%} & 81,66\%         & 81,25\%          & \textit{75,83\%}                   \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
 \caption{Performance of SS-LR LogReg across model parameter choices}
  \label{table_two}
\end{table}

We made several noteworthy observations (Table \ref{table_two}).
%
First, the purely supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall or f1 scores on HCP data.
Classification by low-rank logistic regression is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-dataset performance with small choices for $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
results in more effective feature representations in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-dataset performance
of X and Y, respectively (not shown in Table \ref{table_two}).
This confirms that guiding model learning by task-unrelated structure
extract features of general relevance
beyond the supervised problem at hand.

@@

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the gain of introducing a bottleneck layer
disregarding the autoencoder.
To this end, ordinary logstic regression was juxtaposed
with LR LogRed at $\lambda=1$ (no autoencoder).
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in
generalization perforamnce of
$\approx{17\%}$ (Figure \ref{fig_dimred}).

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.png}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts f1 scores for prediction of each of 38 classes across training epochs.
Ordinary logistic regression operating in voxel space (\textit{left plot})
converged faster but performed worse than
low-rank logistic regression for few
(\textit{middle plot}) and many (\textit{right plot}) latent modes.
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Hence, projecting the input data into a reduced space for classification
yields higher class separability.
}
\label{fig_dimred}
\end{figure}

@@

\paragraph{Feature identification.}
We finally examined whether the models
were fit for purpose
(Figure \ref{fig_weights}).
%
To this end, we computed Pearson's correlation between the classifier weights
and the averaged neural activity map for each of the 18 tasks.
%
Ordinary logistic regression thus yielded a mean feature identification
of $\rho=0.28$.
%
For SS-LR LogReg ($\lambda=0.25, 0.5, 0.75, 1$),
a per-class-weight map was computed by matrix
multiplication of the two inner layers.
Feature identification thus ranged
between $\rho=0.35$ and $\rho=0.55$ for $n=5$,
between $\rho=0.59$ and $\rho=0.69$ for $n=20$, and
between $\rho=0.58$ and $\rho=0.69$ for $n=100$.
%
Consequently,
SS-LR LogReg puts higher absolute weights on relevant structure.
This reflect a higher signal-to-noise ratio, in part explained
by the more BOLD-typical local contiguity.
%
Conversely, SS-LR LogReg puts lower probability mass on irrelevant structure.
Despite lower interpretability, the
salt-and-pepper-like weight maps obtained from
ordinary logistic regression are sufficient for
good classification performance.
%
Hence, SS-LR LogReg yielded class weights 
that were much more similar to features of the respective training samples
for any choice of $\lambda$.
SS-LR LogReg therefore
captures genuine properties of neural activity patterns,
rather than participant- or study-specific artefacts.

\begin{figure}
\begin{centering}
\includegraphics[width=0.98\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\vspace{-0.1cm}
\caption{\textbf{Classification weight maps}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised low-rank logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and thresholded at the 75th percentile.
}
\label{fig_weights}
\end{figure}

\paragraph{Miscellaneous observations.}
For the sake of completeness,
we informally report modifications of the model
that did not improve the generalization performance.
%
Introducing stochasticity into model learning by 
a) input corruption of $\mathbf{X_{task}}$ or
b) drop-out at the bottleneck $\mathbf{W_0}$ using binomial
distributions ($p=0.1, 0.3, 0.5$) deteriorated model performance
in all instances. 
%
Adding c) rectified linear units (ReLU) to $\mathbf{W_0}$ or
other commonly used nonlinearities (d) sigmoid, e) softplus,
f) hyperbolic tangent) all led to smaller classification accuracies.
%
Further, g) ``pretraining'' of the bottleneck $\mathbf{W_0}$
(i.e., non-random initialization) by 
either corresponding SPCA or ICA loadings did not exhibit improved accuracies,
neither did h) autoencoder pretraining.
%
Moreover,
introducing an additional i) overcomplete layer (100 units)
after the bottleneck was not advantageous.
%
Finally, imposing imposing either j) only $\ell_1$-penalty or 
k) only $\ell_2$-penalty was disadvantageous in all tested cases,
which favored ElasticNet regularization chosen in the above analyses.

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of autoencoders, we
learn the optimal decomposition from high-dimensional
voxel brain space into the most important activity patterns for a
supervised learning question.
The higher generalization accuracy and feature recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
This domain-specifc classification algorithms
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
neural activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns. That is,
fMRI data probably concentrate near
a low-dimensional manifold of brain networks.
Extracting fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
We hope that these first steps stimulate development towards
powerful semi-supervised classification methods in systems neuroscience.

@@

\paragraph{Acknowledgment}
{\small
Data were provided by the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.).
}

>>>>

The labeled and unlabeled data were fed into a linear statistical model
composed of an autoencoder and low-rank logistic regression.

XXX (Bertrand): You can simplify data description if needed for the sake of place. It's OK as long as you can refer to a more complete publication describing the data.

\paragraph{Linear autoencoder.}
The affine autoencoder takes the input 
$\mathbf{x}$ and projects it into coordinates of a
latent representation $\mathbf{z}$
and reconstructs it back to $\mathbf{x'}$ by

@@

\begin{equation}
  \mathbb E\left[{\mathcal{L_{AE}}}(\mathbf{x})\right] = %
  \mathbb E\left[\| \mathbf{x} - \mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}) \|^2\right],
\end{equation}
%with \(f_{AE}(\mathbf x) = %
%\mathbf{W_1}(\mathbf{W_0}\mathbf{x} + \mathbf{b_0}) + \mathbf{b_1}\)
% the autoencoder.
% XXX Michael: I don't know to what extent the following makes sense or is
% XXX correct, so I am commenting it
% XXX Danilo: I found this in Bengio's papers somewhere
This reconstruction error criterion equates with
maximizing a lower bound on the mutual information between
input and the learned representation.
Here we choose $\mathbf{W_0}$ and $\mathbf{W_1}$ to be tied, i.e.
$\mathbf{W_0} = \mathbf{W_1^T}$. In particular, this means that the weights
learned are forced to take a two-fold function: That of signal 
\textit{analysis} and that of signal \textit{synthesis}.
The first layer \textit{analyzes}
the data in order to obtain the cleanest latent representation,
while the second
layer represents building blocks from which to \textit{synthesize} the data
using the latent activations.
Tying these processes together makes the analysis
layer interpretable and pulls all non-zero singular values towards 1. 
Nonlinearities were not applied to the
activations in the first layer.

@@

\paragraph{Optimization.}
The common objective was optimized
in the neuroimaging data
by gradient descent in the parameters
of the 
semi-supervised low-rank logistic regression (SS-LR LogReg).
The required gradients are easily obtained by using the chain rule to
backpropagate error derivatives through the linear network.
As solver, we chose \textit{rmsprop} \cite{rmsprop}.
%, a mini-batch version of rprop.
This procedure dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
Gradient normalization by \textit{rmsprop}
is known to effectively exploit curvature information.
We opted for a small batch size of $100$, given the high degree of
redundancy in $X_{rest}$ and $X_{task}$.
The matrix parameters were initalized by Gaussian random values multiplied
by a gain of $0.004$. 
%

XXX: Sorry, I don't like magic numbers. Please explain where this comes from
 ? Maybe $\sqrt{1 / p}$ ?

%
The bias parameters were initalized to zero.
With a slight abuse of notation, let $\theta$ denote a component of $\theta$.
The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch

@@

% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ??
where $0 < \rho < 1$ constitutes the decay rate. $\rho$ was set to
$0.9$ to deemphasize the magnitude of the gradient.
Further, $\alpha$ is the learning rate and $\epsilon$ a global damping factor.
The hyper-parameter $\alpha$ was set to $0.00001$ by prior studies and $\epsilon$ was set to $10^{-6}$.
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with higher generalization performance.

@@

\section{Experimental Results}
\paragraph{Serial versus parallel structure discovery and classification.}
We first tested for an advantage of combined unsupervised decomposition
and supervised classification learning.
We benchmarked against
performing data reduction on the (unlabeled)
first half of the HCP data by ICA and SPCA ($n=5, 20, 50, 100$ components)
and learning classification models in the (labeled) second half
by ordinary logistic regression.
%
ICA performed iterative blind source separation
by a parallel FASTICA implementation (200 maximum iterations,
per-iteration tolerance of 0.0001,
initialized by random mixing matrix, whitening).
SPCA separated the BOLD signals into
network components with few regions by
a regression-type optimization problem constrained by
$\ell_1$-penalty
(no orthogonality assumptions, 1000 maximum iterations,
per-iteration tolerance of 1 * 10\textsuperscript{-8}, sparsity
alpha=1).
%
The second half of the data was projected onto the
latent components discovered in first data half.
The ensuing component loadings were submitted to ordinary
logistic regression
(one hidden layer, $\ell_1=0.1$, $\ell_2=0.1$, 500 maximum iterations).
%
This serail two-step approach was
compared against SS-LR LogReg
(two hidden layers, $\ell_1=0.1$, $\ell_2=0.1$,
500 maximum iterations, $\lambda=0.75$).
Importantly, all trained classification models were tested
on a large, unseen test set (20\% of data) in the presented analyses.
%
Across choices for $n$, SS-LR LogReg
achieved more than 96\% out-of-sample accuracy, whereas
supervised learning based on ICA and SPCA loadings
ranged from 38\% to 87\%
and
32\% to 84\%, respectively (Table \ref{table_one}).
%
These explorations attest to the advantage of directly searching for
classification-relevant structure in the fMRI data,
rather than solving supervised and unsupervised problems independently.
This effect was particularly pronounced when assuming few hidden dimensions.

\begin{table}[h]
 \centering
 \resizebox{0.5\textwidth}{!}{%
 \begin{tabular}{lccc}
 \hline
 \textit{n}               & ICA + LogReg & SPCA + LogReg & SS-LR LogReg \\ \hline
 \multicolumn{1}{l|}{5}   & 37,53 \%     & 32,19 \%      & 96,50 \%     \\
 \multicolumn{1}{l|}{20}  & 80,98 \%     & 78,15 \%      & 97,33 \%     \\
 \multicolumn{1}{l|}{50}  & 84,19 \%     & 83,97 \%      & 97,69 \%     \\
 \multicolumn{1}{l|}{100} & 87,28 \%     & 82,19 \%      & 97,80 \%     \\ \hline
 \end{tabular}
 }
 \vspace{-0.2cm}
 \caption{Serial versus parallel dimensionality reduction and classification
XXX: only one decimal is meaningful}
 \label{table_one}
 \end{table}

@@

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\begin{tabular}{l|cccc|cccc|cccc|cccc|}
  \hline
                            & \multicolumn{4}{c|}{\textit{n = 5}}                                                                                                            & \multicolumn{4}{c|}{\textit{n = 20}}                                                                                                                   & \multicolumn{4}{c}{\textit{n = 100}}                                                       \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.5$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1$} & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.5$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1$}   & $\lambda=0.25$   & $\lambda=0.5$    & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1$}    \\ \hline
  Out-of-sample\\accuracy   & \textit{88,90\%}                   & 95,13\%                            & \textbf{96,49\%}                  & 95,72\%                          & 97,44\%                              & \textbf{97,80\%}                    & 97,33\%                              & \textit{97,32\%}                   & 97,21\%          & \textit{97,03\%} & \textbf{97,80\%} & 97,38\%                            \\
  Precision (mean)          & \textit{86,98\%}                   & 94,86\%                            & \textbf{96,28\%}                  & 95,36\%                          & \textbf{97,38\%}                     & 97,06\%                             & \textit{96,98\%}                     & 97,02\%                            & 96,90\%          & \textit{96,47\%} & \textbf{97,50\%} & 96,91\%                            \\
  Recall (mean)             & \textit{88,27\%}                   & 95,19\%                            & \textbf{96,57\%}                  & 95,70\%                          & \textbf{97,51\%}                     & 97,50\%                             & \textit{97,35\%}                     & 97,36\%                            & 97,20\%          & \textit{97,15\%} & \textbf{97,88\%} & 97,37\%                            \\
  F1 score (mean)           & \textit{86,64\%}                   & 94,89\%                            & \textbf{96,38\%}                  & 95,43\%                          & \textbf{97,42\%}                     & 97,22\%                             & \textit{97,11\%}                     & 97,13\%                            & 97,00\%          & \textit{96,71\%} & \textbf{97,65\%} & 97,07\%                            \\ \hline
  Out-of-dataset \\accuracy & 60,83\%                            & \textit{54,30\%}                   & 60,69\%                           & \textbf{62,92\%}                 & 79,72\%                              & \textbf{81,94\%} & 79,72\%          & \textit{79,44\%}                     & \textbf{82,08\%} & 81,66\%         & 81,25\%          & \textit{75,83\%}                   \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
 \caption{Performance of SS-LR LogReg across model parameter choices. XXX: recall what the chance-level performance is.XXX: please add lambda=0}
  \label{table_two}
\end{table}

We made several noteworthy observations (Table \ref{table_two}).
%
First, the purely supervised estimator ($\lambda=1$) achieved in no
instance the best accuracy, precision, recall or f1 scores on HCP data.
Classification by low-rank logistic regression is therefore facilitated by
imposing structure from the unlabeled rest data.
%
Second, the higher the number of latent components $n$,
the higher the out-of-dataset performance with small choices for $\lambda$.
This suggests that the presence of more rest-data-inspired hidden components
results in more effective feature representations in unrelated task data.
%
Third, for $n=20$ and $100$ (but not $5$) the purely rest-data-trained
decomposition matrix ($\lambda=0$) resulted in
noninferior out-of-dataset performance
of X and Y, respectively (not shown in Table \ref{table_two} XXX: please show it).
This confirms that guiding model learning by task-unrelated structure
extract features of general relevance
beyond the supervised problem at hand.

@@

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the gain of introducing a bottleneck layer
disregarding the autoencoder.
To this end, ordinary logstic regression was juxtaposed
with LR LogReg  at $\lambda=1$ (no autoencoder).
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in
generalization perforamnce of
$\approx{17\%}$ (Figure \ref{fig_dimred}).

\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.png}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts f1 scores for prediction of each of 38 classes across training epochs.
Ordinary logistic regression operating in voxel space (\textit{left plot})
converged faster but performed worse than
low-rank logistic regression for few
(\textit{middle plot}) and many (\textit{right plot}) latent modes.
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Hence, projecting the input data into a reduced space for classification
yields higher class separability.
XXX: ugly figure: use larger font for labels and titles.
XXX: includes pdf figures, not a png file.
}
\label{fig_dimred}
\end{figure}

@@

\paragraph{Feature identification.}
We finally examined whether the models
were fit for purpose
(Figure \ref{fig_weights}).
%
To this end, we computed Pearson's correlation between the classifier weights
and the averaged neural activity map for each of the 18 tasks.
%
Ordinary logistic regression thus yielded a mean correlation of
of $\rho=0.28$.
%
For SS-LR LogReg ($\lambda=0.25, 0.5, 0.75, 1$),
a per-class-weight map was computed by matrix
multiplication of the two inner layers.
Feature identification thus ranged
between $\rho=0.35$ and $\rho=0.55$ for $n=5$,
between $\rho=0.59$ and $\rho=0.69$ for $n=20$, and
between $\rho=0.58$ and $\rho=0.69$ for $n=100$.
%
Consequently,
SS-LR LogReg puts higher absolute weights on relevant structure.
This reflect a higher signal-to-noise ratio, in part explained
by the more BOLD-typical local contiguity.
%
Conversely, SS-LR LogReg puts lower probability mass on irrelevant structure.
Despite lower interpretability, the
salt-and-pepper-like weight maps obtained from
ordinary logistic regression are sufficient for
good classification performance.
%
Hence, SS-LR LogReg yielded class weights 
that were much more similar to features of the respective training samples
for any choice of $\lambda$.
SS-LR LogReg therefore
captures genuine properties of neural activity patterns,
rather than participant- or study-specific artefacts.
XXX: But what you show here is mostly a denoising or regularization effect.

\begin{figure}
\begin{centering}
\includegraphics[width=0.98\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\vspace{-0.1cm}
\caption{\textbf{Classification weight maps}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} ordinary logistic regression (same
implementation but without bottleneck or autoencoder),
XXX: sorry, but is this a multinomial or plain logistic ? Please clarify 
\textit{middle column:} semi-supervised low-rank logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Low-rank logistic regression
a) puts higher absolute weights on relevant structure
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity.
All values are z-scored and thresholded at the 75th percentile.
}
\label{fig_weights}
\end{figure}

\paragraph{Miscellaneous observations.}
For the sake of completeness,
we informally report modifications of the model
that did not improve the generalization performance.
%
Introducing stochasticity into model learning by 
a) input corruption of $\mathbf{X_{task}}$ or
b) drop-out at the bottleneck $\mathbf{W_0}$ using binomial
distributions ($p=0.1, 0.3, 0.5$) deteriorated model performance
in all instances. 
%
Adding c) rectified linear units (ReLU) to $\mathbf{W_0}$ or
other commonly used nonlinearities (d) sigmoid, e) softplus,
f) hyperbolic tangent) all led to smaller classification accuracies.
%
\textbf{XXX: Please give explanations why this is the case. My feeling is that we're still not rich enough in the sample dimension to make these tricks worthwhile.}
Further, g) ``pretraining'' of the bottleneck $\mathbf{W_0}$
(i.e., non-random initialization) by 
either corresponding SPCA or ICA loadings did not exhibit improved accuracies,
neither did h) autoencoder pretraining.
%
Moreover,
introducing an additional i) overcomplete layer (100 units)
after the bottleneck was not advantageous.
%
Finally, imposing imposing either j) only $\ell_1$-penalty or 
k) only $\ell_2$-penalty was disadvantageous in all tested cases,
which favored ElasticNet regularization chosen in the above analyses.

\section{Discussion and Conclusion}
% Conceptual recap+gain of current paper
Using the flexibility of autoencoders, we
learn the optimal decomposition from high-dimensional
voxel brain space into the most important activity patterns for a
supervised learning question.
The higher generalization accuracy and feature recovery, comparing to
ordinary logistic regression, hold potential
for adoption in various neuroimaging analyses.
Besides increased performance, these models are more interpretable by
automatically learning a mapping to and from a brain-network space.
This domain-specifc classification algorithms
encourages departure from the artificial and statistically
less attractive voxel space.
Neurobiologically,
neural activity underlying defined mental operations
can be explained by linear combinations of the main activity
patterns. That is,
fMRI data probably concentrate near
a low-dimensional manifold that correspond to peculiar brain networks combinations.
Extracting fundamental building blocks of brain organization might
facilitate the quest for the cognitive primitives of
human thought.
We hope that these first steps stimulate development towards
powerful semi-supervised classification methods in systems neuroscience.

@@

\paragraph{Acknowledgment}
{\small
Data were provided by the Human Connectome Project. The study was supported
by the German National Academic Foundation (D.B.).
XXX: Note that this will be removed at submission for the sake of anonymization.
}
